Opening up the court (surface) in tennis grand slams

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
merge

## Data {#sec:data}
The data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).  Each match has `r ncol(gs)` attributes, many of which are redundant.  We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.  Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.  A subset of the data is shown in Table \ref{tab:data}.

@@

The data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.

>>>>

## Data {#sec:data}

The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).  Each match has `r ncol(gs)` attributes, many of which are redundant.  We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.  Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.  A subset of the data is shown in Table \ref{tab:data}.

@@

The secondary data point by point data for grand slam matches.  In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.  There is also tournament info such as court, year, and time start.  Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.  However, these additional attributes are only available for about 1/10 of the data.

We aggregate the point by point data into match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.  We then join the point by point match data with the grand slam data to find the score of the match.  We do this by matching on tournament, year, player one, and player two.  This final data set consists of 3066 observations (compared to the 5080 in the original grand slams).  The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)` and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.  The same trend is true for the loser rank.  This means that it is more likely for better players to have court by court data, which may effect our analysis.  Additionally, we found that no women's point by point data was recorded for 2015.

All data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.

### 2018-09-15 contest/surface-type/paper-tennis.tex
merge

The data consists of 5080 matches split evenly over the four grand slams
and the two leagues: ATP (men's) and WTA (women's). Each match has 80
attributes, many of which are redundant. We focus on the following
attributes for both the winner and loser of the match: games won, points
won, retirement, break points faced, break points saved, aces, country
of origin, and player attributes. Additionally, we take into account the
number of sets in a match, the surface type, and round of the
tournament. A subset of the data is shown in Table \ref{tab:data}.

@@

The data is obtained from Jeff Sackmann's open website via the
\texttt{R} package \texttt{deuce} (Sackmann 2018; Kovalchik 2017). All
steps of our analysis from collection to dissemination are freely
available online.

@@

>>>>

The primary data consists of 5080 matches split evenly over the four
grand slams and the two leagues: ATP (men's) and WTA (women's). Each
match has 80 attributes, many of which are redundant. We focus on the
following attributes for both the winner and loser of the match: games
won, points won, retirement, break points faced, break points saved,
aces, country of origin, and player attributes. Additionally, we take
into account the number of sets in a match, the surface type, and round
of the tournament. A subset of the data is shown in Table
\ref{tab:data}.

@@

The secondary data point by point data for grand slam matches. In this
data set, each row is a point in a match with details on who won the
point, whether a player had a forced or unforced error, winner, ace, or
net point win, and serve speed. There is also tournament info such as
court, year, and time start. Additional attributes include rally length,
winner and final score of the match, retirement, and minutes played.
However, these additional attributes are only available for about 1/10
of the data.

We aggregate the point by point data into match data, summing the total
number of errors, winners, aces, net points, and taking the average
service speed for each player in the match. We then join the point by
point match data with the grand slam data to find the score of the
match. We do this by matching on tournament, year, player one, and
player two. This final data set consists of 3066 observations (compared
to the 5080 in the original grand slams). The median rank of the winners
for the original data is 28 and the median rank for the winners of the
partial data is 21. The same trend is true for the loser rank. This
means that it is more likely for better players to have court by court
data, which may effect our analysis. Additionally, we found that no
women's point by point data was recorded for 2015.

All data is obtained from Jeff Sackmann's open website via the
\texttt{R} package \texttt{deuce} (Sackmann 2018; Kovalchik 2017). All
steps of our analysis from collection to dissemination are freely
available online.

@@

\hypertarget{mixed-effects-models-for-all-grand-slams}{%
\subsection{Mixed-effects models for all grand
slams}\label{mixed-effects-models-for-all-grand-slams}}

\hypertarget{logistic-regression-for-wins}{%
\subsubsection{Logistic regression for
wins}\label{logistic-regression-for-wins}}

We will use a multilevel logistic regression framework. This allows us
to include fixed effects, which remain the same for each player, as well
as include player-level effects. Since players appear multiple times in
the data, the observations are not independent and including a
player-level effect allows us to account for this dependence. It also
provides a way to examine individual player effects and assess, for
instance, whether Federer is more likely to win at Wimbledon after
accounting for other variables.

We initially planned to use points won by an individual as our outcome
variable, but found that it may not be an ideal measure of player
performance. For instance, players may score few points in a match due
to a poor performance, but they may also score few (relative) points if
they win a short match. Modeling the number of points won is also
complicated by the difference between length of matches for men (best of
5 sets) and women (best of 3 sets). For these reasons, we chose to model
whether a player won the match, with the following predictors:

\begin{itemize}
\tightlist
\item
  \texttt{ioc\_fac}: Country that the player represents
\item
  \texttt{tournament}: Which Grand Slam the match was played at
\item
  \texttt{late\_round}: Indicator for whether the match occured in Round
  of 16 or later
\item
  \texttt{rank}: Rank of player at the time of the match

  \begin{itemize}
  \tightlist
  \item
    Included on log scale
  \end{itemize}
\item
  \texttt{opponent\_rank}: Rank of opponent at the time of the match

  \begin{itemize}
  \tightlist
  \item
    Included on log scale
  \end{itemize}
\item
  \texttt{year}: Factor variable with a level for each year included in
  the dataset (2013-2017)
\item
  \texttt{atp}: Indicator that the match was played in the ATP league
  instead of the WTF league.
\end{itemize}

(short discussion of models fit, show plots for final model)

\hypertarget{linear-regression-for-other-stats}{%
\subsubsection{Linear regression for other
stats}\label{linear-regression-for-other-stats}}

Rank and opponent's rank clearly have the largest effect on predicting
the winner. Looking at the individual effects for Serena, Nadal, and
Federer; they actually fall near the average for all four grand slams.
Including rank as an indicator (seeded/not-seeded) substantially
decreases the fit of the model, even when individual effects are
included. We now look at a similar model, but instead of using `win' as
the outcome variable, we'll look at game statistics that may be more
sensitive to court surface -- aces, winners, and net points won -- and
see if individual differences are detectable.

(some plots that show a) Serena/Federer are stronger servers, b) Federer
makes fewer unforced errors at Wimbledon, c) something about Nadal)
